An open population age structured distance sampling model

### 2023-06-01 Manuscript_Nilsen_Nater.qmd
Embedded figures in manuscript
For now, I've just added all figures that may be relevant. Some may be moved to supplementary. Figures from simulated data tests still to follow. 

Note that I have not properly set up the referencing yet. We should also look into automatic numbering of figures.

### Field data

Line transect survey data were collected in August each year, prior to the annual autumn harvest season. Line transects are surveyed by educated volunteers, using pointing dogs to located the birds. When located, the geographical coordinate, perpendicular distance from the sampling line, the number of birds as well as the age (juvenile and adult) and sex of the birds are recorded. Additional information about the surveys are found in Nilsen *et al.* (2020) and Bowler *et al.* (2020). \[Add sample sizes\].

@@

**Figure 1.** Life cycle graph

We assumed that the population size and structured was surveyed each year at the end of the summer season in August, equivalent to the schedule of the real survey data. Ptarmigan breed annually, and the clutch hatch around last week of June. By the time of survey, they are around 1.5 months of age during the annual line transect surveys in August. Ptarmigan reach maturity at the age of one (ref). The life cycle is therefore described in term of i) annual survival probability (S~t~) which describe the probability to survive from August in year *t* to August in year *t+1*. Here, we assume no age-or sex-specificity in the survival rates, acknowledging that this is a simplifying assumption that could be alleviated with additional auxiliary data (Sandercock *et al.* 2011; Israelsen *et al.* 2020). Moreover, following the standard for post-breeding census models year specific recruitment (R~t~) will be given by the product of annual survival (S~t~) and fecundity (F~t~):

(equation 1)

@@

>>>>

@@

![**Figure 1.** Simplified graphical representation of the ptarmigan life cycle with two age classes and the data sources included in the integrated distance sampling model. Juv\[t\] = juveniles in year t. Ad\[t\] = adults in year t. R\[t\] = recruitment rate in year t. S\[t\] = survival probability from year t to t+1.](Figures/RypeIDSM_structure.png)

We assumed that the population size and structured was surveyed each year at the end of the summer season in August, equivalent to the schedule of the real survey data. Ptarmigan breed annually, and the clutch hatch around last week of June. By the time of survey, they are around 1.5 months of age during the annual line transect surveys in August. Ptarmigan reach maturity at the age of one (ref). The life cycle is therefore described in term of i) annual survival probability (S~t~) which describe the probability to survive from August in year *t* to August in year *t+1*. Here, we assume no age-or sex-specificity in the survival rates, acknowledging that this is a simplifying assumption that could be alleviated with additional auxiliary data (Sandercock *et al.* 2011; Israelsen *et al.* 2020). Moreover, following the standard for post-breeding census models year specific recruitment (R~t~) will be given by the product of annual survival (S~t~) and fecundity (F~t~).

@@

### Field data (case study)

Line transect survey data were collected in August each year, prior to the annual autumn harvest season. Line transects are surveyed by educated volunteers, using pointing dogs to located the birds. When located, the geographical coordinate, perpendicular distance from the sampling line, the number of birds as well as the age (juvenile and adult) and sex of the birds are recorded. Additional information about the surveys are found in Nilsen *et al.* (2020) and Bowler *et al.* (2020). \[Add sample sizes and additional information on data processing/storage post collection, i.e. how do volunteers report data, storage in hønsefuglportalen, quality control, pushes to LivingNorway and GBIF\].

\[Figure 2: Map of Lierne study area?\]

## Results

### Model performance on simulated datasets

\[One or several figures of estimates vs. true values?\]

### Case study on willow ptarmigans in Lierne

#### Population density

![**Figure X.** Estimated density of willow ptarmigans in Lierne from 2007 to 2021. Solid line represents the posterior median, ribbon marks 95% credible interval.](Figures/TimeSeries_popDens.png)

#### Vital rates

![**Figure X.** Posterior densities of A) seasonal survival, B) average annual survival, and C) recruitment rate. For the latter, the yellow distribution is for the intercept, representing a baseline recruitment rate when rodent occupancy is low. The turquoise distributions are for year-specific estimates or recruitment rate, with darker colors indicating later years.](Figures/PostDens_Mu_S_tR.png)

![**Figure X.** Estimated recruitment rate of willow ptarmigans in Lierne from 2007 to 2021. Solid line represents the posterior median, ribbon marks the 95% credible interval.](Figures/TimeSeries_rRep.png)

![**Figure X.** Predicted effect of rodent occupancy on recruitment rate. Solid line represent the posterior median, ribbon marks the 95% credible interval.](Figures/Rep_betaR.R.png)

#### Detection parameters

![**Figure X.** Predicted average detection probabilities for distance sampling surveys in Lierne from 2007 to 2021. Solid line represents the posterior median, ribbon marks the 95% credible interval.](Figures/TimeSeries_pDetect.png)

## Discussion

### Conclusion

### Author contributions

### Acknowledgements

### Data and code availability
